Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-12 ThesisJP/Background/background.tex
Finished review from CI.

This chapter gives the background information needed in order to understand the topics of this thesis and its papers. First, in Section \ref{sec:gpgpu}, the history and the principle of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) are reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}. Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate until the required cooling finally became to much in 2004.  From that point, CPU manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and colors where fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.

The graphics pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advanced feature found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to unify the computational power of both processors. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers. An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU. This is a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/CPU-GPU.pdf}
\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architecture. Note how more space is dedicated to compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
\label{fig:cpu_gpu}
\end{figure}

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture by Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15*192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.}
\label{fig:cpu_vs_gpu}
\end{figure}

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the graphics pipeline. When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

\begin{figure}
\centering
\subfloat[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfloat[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square.}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which values to gather from the global memory, in this case an element of \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in the block. Thus,  in order to formulate a given problem as a kernel function, the global memory input to each thread must not depend on the output from other blocks. Hence, it must be possible for the GPU to schedule blocks in any order.

@@

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into enough independent calculations. Also, to achieve the peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution. The geometry of cardiac ultrasound imaging is depicted in Fig.\,\ref{fig:us_grid}. A line in an ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. When transmitting the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}, where $f_c$ is the center frequency of the transmitted pulse. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.

@@

In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes. This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool). The added complexity of this technique caused by the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.

>>>>

This chapter gives the background information needed in order to understand the topics of this thesis and the associated papers. First, in Section \ref{sec:gpgpu}, the history and the principles of general-purpose computing on graphics processing units (\nom{GPGPU}{General-purpose computing on graphics processing units}) are reviewed, followed by an introduction to medical ultrasound imaging and adaptive beamforming in Section \ref{sec:ultrasound}. Finally, the concepts of volume rendering and ultrasound field simulations are presented in Sections \ref{sec:volren} and \ref{sec:field}.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry in a recent essay \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spin around the fact that manufacturers of central processing units (CPUs) hit a frequency wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single CPU \textit{processing core}, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation were also growing at the same rate until the required cooling finally became too much in 2004.  From that point, CPU manufactures had to pursue different directions than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to annually gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

@@

Around the same time as CPU manufacturers hit the frequency wall, people had started experimenting with programmable shaders, which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and colors were fed to the graphics processing unit (GPU) in one end, and where shaded and rasterized sceneries were outputted in the other. Programmable shaders introduced the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this development was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall, the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games.

The original graphics pipeline involved no conditional branching, and data were typically used once \cite{Purcell2002}. Designers of GPUs could therefore skip advanced features found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to the GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_gpu} and \ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the problem of rendering graphics. When geometry is shaded, projected, and rasterized, the same instruction is needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction type)}) \cite{Flynn1966}, especially suited for this problem. How SIMD instructions are utilized in modern CPUs and GPUs is discussed in Section \ref{sec:cpu_vs_gpu}.

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to get the most out of each processors. Balancing means that highly parallel and computationally intense tasks should be offloaded to the GPU while serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as \textit{heterogeneous computing}. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array (processor)}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processor is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics, and all of them are therefore APUs. An on-chip GPU is less powerful than a high-performance discrete GPU, but valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the CPU. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers. An integrated GPU can therefore provide increased performance over a discrete GPU if the problem requires a high level of CPU-GPU data synchronization.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-cores, heterogeneous cores, and cloud cores form the "jungle" that today's software engineers have to handle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with a multi-core CPU and a discrete high performance GPU. This is a cheap, high-performance system which is likely to be the specification of current and future ultrasound imaging system.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/CPU-GPU.pdf}
\caption{Schematic overview of CPU (Intel Haswell) and GPU (Nvidia Kepler) architectures. Note how more space is dedicated to compute cores on a GPU, and how more space is dedicated to control and caching on a CPU.}
\label{fig:cpu_gpu}
\end{figure}

In Section \ref{sec:gpgpu} a special type of instructions was mentioned, namely SIMD instructions. These are instructions which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction set)}). The first SSE instruction set had registers of 128 bit which meant that four single-precision (32 bit) floating point values could be processed per clock cycle. Updates of the SSE instruction set (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction set)}, AVX2, and AVX-512) have added new instructions and longer registers. The newly introduced extensions, AVX and AVX2, add registers of 256 bit and fused multiply accumulate (\nom{FMA}{Fused multiply-accumulate (instruction)}) instructions  which means that sixteen single-precision floating point operations can be performed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports per core supporting FMA-AVX2 instructions, and is therefore capable of performing 32 single-precision floating point operations per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight-core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014. The numbers in Fig.\,\ref{fig:cpu_vs_gpu} for the last three generations of CPUs are all based on 8-core CPUs.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32 \times 8 \times 3.8 = 972.8\,\text{GFLOPS}.
\end{equation}
Hence, a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture (Single precision). In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, has 40 execution units capable of processing sixteen single-precision floating point values at 1.3 GHz. This adds
\begin{equation}
16 \times 40 \times 1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as units consisting of multiple SIMD cores (Fig.\,\ref{fig:cpu_gpu}). The recent Kepler architecture from Nvidia can provide up to 15 SIMD cores\footnote{Nvidia likes to refer to the unit capable of processing one single precision floating point value as a core. In Nvidia's world, Kepler therefore has $15\times192=2880$ cores.} capable of processing 192 single-precision floating point values per clock cycle per core. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The maximum theoretical peak single precision FLOPS for this architecture with a core frequency of 745 MHz is then found to be
\begin{equation}
15 \times 192 \times 2 \times 745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also includes 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of 5 TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with the SIMD cores on CPUs, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{img/cpu_vs_gpu.pdf}
\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs.}
\label{fig:cpu_vs_gpu}
\end{figure}

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge of computer graphics and the graphics pipeline. When GPGPU gained a broader interest and the GPU started to be used to compute problems that did not involve graphics, it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU framework)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU framework)}, a standard which is maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. While CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

\begin{figure}[t!]
\centering
\subfloat[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfloat[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square.}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

When the kernel is launched, a copy of it is scheduled for each thread in each block, and each kernel instance gets as input its location in the grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which values to gather from the global memory, in this case an element of \texttt{a}. %Behind the scene, thread blocks are divided into smaller sets of threads called a warp (or wavefront in OpenCL). 
When a kernel is executed, multiple arbitrary thread blocks are scheduled to each SIMD core where they stay until all instructions in the kernel are finished for all threads in all the blocks. This step is then repeated until all blocks have been processed. Thus,  in order to formulate a given problem as a kernel function, it must be possible for the GPU to schedule the resulting thread blocks in any order. As an example, the global memory input to each thread can therefore not depend on the output from other blocks.

@@

It is important to understand that not all algorithms will benefit from wider SIMD registers. As the width increase, the only algorithms that will benefit are those which can be divided into a sufficiently large number of independent calculations. Also, to achieve peak performance, the implementation must perform one FMA operation on each core per clock cycle, and it must be possible for the GPU to hide all memory latency by scheduling threads which are not stalled by a memory fetch. All of this is something which is impossible to achieve in practical applications, and therefore one will always end up with a throughput which is a fraction of the peak throughput.

@@

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence, energy will be distributed or sensed equally in all directions. When multiple omnidirectional elements are arranged in an array, the transmitted pulse will get shaped into a beam\footnote{A beam refers to the accumulated magnitude across time of a pulse traveling in space.} of ultrasound by diffraction as the width (the \textit{aperture}) of the array grows. The angular width of this beam, and also the angular resolution in radians of the array, is approximately given by \cite{AngelUltrasound}:
\begin{align}\label{eq:res}
\delta\theta_a = \frac{\lambda}{D} = \frac{c}{fD},
\end{align}
\nomi{$\delta\theta_a$}{Angular resolution of an aperture}where \nom{$\lambda$}{Wavelength} and \nom{$f$}{Frequency} are the ultrasound wavelength and frequency respectively, $c$ is the speed of sound in human tissue, and \nom{$D$}{Aperture size} is the aperture width. With an average speed of sound in human tissue of $1540\frac{\text{m}}{\text{s}}$, the angular resolution or opening angle of a 2 cm cardiac probe operating at 3 MHz is around 1.5\degree. The axial resolution is inversely proportional to the pulse length, and is typically better than the lateral resolution. An overall view of the cardiac ultrasound imaging geometry is depicted in Fig.\,\ref{fig:us_grid}. A line in the depicted ultrasound image is generated by firing an ultrasound pulse or beam into the body and sample the received echo. We will refer to this amplitude trace as an \textit{image line}.

@@

The standard form of beamforming used in ultrasound imaging is known as \textit{delay-and-sum} (DAS): 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\{\Delta_m[n]\}$.  From now on we will omit the dependency on $\{\Delta_m[n]\}$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. When transmitting the pipeline depicted in Fig.\,\ref{fig:das_background}, without the sum operator, is reversed and the negated set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

The required axial sampling is dictated by the transmit frequency and the frequency bandwidth of the probe. From Fourier theory we know that a short pulse consists of many frequency components. A short ultrasound pulse, which would yield a high axial resolution, requires therefore a large frequency bandwidth. Medical ultrasound systems typically has a bandwidth between 33 \% and 100 \% of the pulse frequency. This corresponds to transmitting a pulse which is between 1 and 3 oscillations when entering tissue. In the analog-to-digital converter of an ultrasound imaging system a high sampling frequency is typically used (e.g. 40 MHz). However, the ultrasound signal is band limited, and the data rate can therefore be reduced in a process known as \textit{in-phase quadrature (\nom{IQ}{In-phase quadrature (ultrasound data format).}) demodulation}. In this process the axial sampling rate is reduced to the bandwidth of the imaging system in the following three steps: demodulation with a complex exponential, low-pass filtering, and decimation. The number of samples in one image line is after this process given by:
\begin{align}
N = \frac{2r}{c}B_wf_c,
\end{align}
\nomi{$N$}{Number of samples in range (or time)}where $f_c$ is the center frequency of the transmitted pulse, and where \nom{$B_w$}{Relative bandwidth} is the relative bandwidth in percent of \nom{$f_c$}{Center frequency of ultrasound pulse}. Note that after this process the ultrasound data will be complex values. The total number of samples in an B-mode image from a phased array is then 
\begin{align}
N_{\text{B-mode}} = N_\theta N.
\end{align}
If our cardiac probe, which is used to image a $70$\degree{} sector, has a relative bandwidth of $80$ \% we have to process almost $50000$ complex data vectors of length $M$ every $1/55$ second. The formula for the number of FLOPS required for real-time DAS beamforming of cardiac ultrasound is then roughly given as:
\begin{align}
\text{FLOPS}_{\text{DAS}} = MN_{\text{B-mode}}f_{\text{FPS}} = MB_wf_c.
\end{align}
If our cardiac probe has $M=128$ elements, DAS beamforming requires a processing throughput of around $300$ MFLOPS. This is far below the peak processing throughput of modern GPUs and CPUs. The main challange with DAS beamforming is the amount of data that has to be moved. In our example, for a sample size of four bytes, around 1.2 GB has to be moved to the GPU every second. This is however less than the maximum bandwidth of PCI Express 2.0 (8 GB/s), even after the relative bandwidth, transmit frequency and number of array elements are increased with a total factor of 5. Newer versions of PCI Express, which supports even higher transfer rates, have also recently been introduced.

@@

In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes. This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool). The added complexity of this technique is caused by the statistical estimation, which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.
